Giant Low Surface Brightness Galaxies As Merger Remnants

### 2017-01-24 final/ms.tex
put sentences on own lines
remove 'normal'

\begin{abstract}
We present IFU observations of the giant low surface brightness galaxies Malin~2 and UGC~6614.  We fit the stellar dynamics of these systems and compare their spatially resolved spectra to star formation history models.  We argue that Malin~2 and UGC~6614 are normal elliptical galaxies that have undergone high angular momentum mergers which formed their GLSB disks, making them unique systems for probing the dark matter halos of elliptical systems.  We find the stellar dynamics are consistent with earlier measurements of the GLSB \HI\ disks.  The stellar populations of the GLSB disks are dominated by old metal-poor stars, while the central ellipticals are metal rich.

@@

Observers tend to classify GLSB systems as normal spirals.  \citet{Matthews01b} list Malin~2 as an Scd while \citet{Das13} conclude that GLSB systems are an extreme form of late type spiral galaxies.  Similarly, \citet{Kasparova14} claim that the large dark matter scale of Malin~2 can explain it's features.  Even the Galaxy Zoo project \citep{Willett13}  classifies GLSB galaxies as spirals.  While GLSB systems superficially resemble normal spiral galaxies, they are clearly extreme outliers when their scale lengths are compared to other galaxy samples, being around a factor of 2 larger than the largest normal spirals \citep[Figure~4]{Dalcanton97}.

@@

The other popular formation scenario for GLSB systems is that they are normal galaxies that undergo a high angular momentum minor merger where the LSB disk is the shredded remains of an accreted galaxy.  \citet{Pen06} find that extended disks can be formed via the disruption of dwarf galaxies.  These simulated debris disks have exponential profiles and extend out to 200 kpc from the central galaxy.

Observations of Malin~1 support the accretion model of GLSBs.  \citet{Barth07} analyzes HST imaging of Malin~1 and find that the central object is well described as a normal early-type barred galaxy that was previously unresolved in ground-based imaging. \citet{Barth07} points out that extended disk structures in galaxies such as M31 appear to be a smooth continuation of the inner disk \citep{Ferguson02}, while Malin~1's extended disk is better described as a distinct component, (although \citet{Ibata05} claim M31's extended disk is of accretion origin).

@@

We obtained spatially resolved spectroscopy of our target galaxies using the Mitchell Spectrograph (formerly VIRUS-P)\citep{Hill08} on the 2.7m Harlan J. Smith telescope at McDonald Observatory.  Observations were made during dark time in February, March, and April 2010.  Identical instrument setups were used for all observations.  The Mitchell Spectrograph with the VP-2 IFU bundle used here has a square array of 246 optical fibers which sample the 1.9\arcmin$\times$1.9\arcmin\ field-of-view with a 1/3 filling factor, with each fiber having a diameter of 4.23\arcsec.  The spectrograph images the spectrum of each fiber on a 2048$\times$2048 Fairchild Imaging CCD.

@@

Once the Vaccine routines are completed we make the following additional reductions using standard IRAF routines and custom IDL scripts.  The reduced twilight frames are used to generate a fiber-to-fiber flux correction (similar to a longslit illumination correction).  We use the b-spline procedure described in \citet{Kelson03} to generate background frames.  The corners of the IFU array contain enough empty fibers to measure the sky background in each exposure.  Because we bin fibers together, imperfect sky-subtraction represents a serious source of potential systematic errors.

@@

\begin{figure*}
\epsscale{.4}
\plotone{Color_Magnitu_Diagram.eps}\plotone{Faber_Jackson_Relation.eps}\\
\plotone{Fundamental_Plane.eps}\plotone{mg_sigma.eps}
\epsscale{1}
\caption{
Top Left:  The color-magnitude diagram for 141,000 galaxies of similar redshift and magnitude as Malin~2 and UGC~6614.  Top Right:  The Faber-Jackson relation for 45,700 SDSS galaxies, along with a best-fit line.  Left:  The $r$-band fundamental plane from SDSS along with our GLSB galaxies and best-fit line.  Bottom Right:  The relation between central velocity dispersion and the Mg2 absorption feature.   From the point of view of the SDSS data reduction pipeline, GLSB galaxies appear as relatively normal red elliptical galaxies.  \label{elliptical_relations}}
\end{figure*}

@@

The diameters of GLSB disks are very large, with Malin~1's stellar disk detected out to a radius of 80 kpc \citep{Galaz15}.  Stellar disks are typically embedded in much larger \HI\ disks.  In Figure~\ref{disk_comp}, we compare the sizes of the stellar and \HI\ disks of the GLSB galaxies with other observations.  For this plot, we take the maximum detected radius from the literature for the galaxies.  It is possible that deeper observations could increase the values of $R_{max}$.  While the stellar disk of UGC~6614 is close in size to local observed \HI\ disks, Malin~1 and~2 are larger than the \HI\ disks observed in the THINGS survey.  This seems to point against the GLSB formation mechanism of \citet{Mapelli08}, as we do not observe \HI\ disks in normal spirals large enough to convert into GLSB systems unless an interaction is able to also boost the angular momentum of the resulting stellar disk.

@@

As we have shown in \S\ref{sec:centers} and \S\ref{sec:disks}, GLSB galaxies can be well described as large normal ellipticals that have managed to accrete a high angular momentum disk of stars and gas.  Given the total \HI\ mass in GLSBs, the accreted object is consistent with a much lower-mass gas-rich spiral.  If this is in fact the way GLSBs form, the disk kinematics represent a unique probe of the dark matter halo of elliptical galaxies.

@@

\citet{Murphy2011} are able to use IFU observations of M87 to reach $\sim2 R_e$ using just stellar data, and then rely on globular cluster kinematics to extended out to $5 R_e$ where they find a dark matter fraction of 85\%.  \citet{Weij09} is similarly able to use IFU observations to reach $\sim 4 R_e$ on two elliptical galaxies while \citet{Proctor09} reaches $\sim 3 R_e$ using multi-object spectroscopy.

@@

We find that the centers of these two GLSB systems very closely resemble normal galaxies, with the extended faint disks being distinct components, consistent with the previous results for Malin~1 \citep{Barth07} and NGC~7589 \citep{Lelli10}.

@@

It also seems unlikely that GLSBs are just extreme versions of normal disk galaxies \citep{Das13}.  It's hard to imagine how such a large massive central spheroid could be formed without also disrupting the surrounding disk.  While the old stellar populations in the disks of Malin~2 and UGC~6614 are evidence against the \citet{Mapelli08} formation scenario, we have not addressed what collisional ring galaxies do evolve into.  Since photometric surveys are often surface brightness limited, it is possible a GLSB system will be found in the future that is the descendant of a collisional ring system.

One of the most important features we can see in our data is the sharp metallicity gradient as one transitions between the central region and the disk in both Malin~2 and UGC~6614 (Figures~\ref{metal_malin2} and~\ref{metal_ugc}).  For both galaxies, the best fitting star formation histories show the stellar populations are dominated by old stars inside and outside the transition, but the metallicity changes by $\sim0.7-1.0$ dex.  This is a very sharp gradient compared to what has been found in normal spirals for the bulge-to-disk transition region.  \citet{MacArthur09} spectroscopically observed a range of spiral galaxies and found the metallicity profiles which were dropping, rising, and flat.  Using broadband colors, \citet{MacArthur04} also found metallicity gradients were generally very shallow, with steep metallicity gradients only in very late-type spiral galaxies. Also note that our metallicity gradient could be an underestimate of the true gradient since the central metallicities are at the limits of our model grid (i.e., they could have metallicities greater than [Z/H]=0.7). We have also spatially binned data, possibly diluting how sharp the transition is between bulge and disk.

@@

We have observed two giant low surface brightness galaxies with the VIRUS-P IFU.  We find the systems are best explained as massive elliptical galaxies which have managed to accrete stars and gas into a large disk.  We find the inner bulge is similar to elliptical galaxies found in SDSS.  The extended galaxy stars have disk-like velocity fields, are dominated by old stars, but do show signs of ongoing star formation. The stars also show a sharp metallicity transition, consistent with the disk forming independently of the bulge. The \HI\ mass of the disk is also consistent with coming from a low-mass accreted galaxy.

We are currently observing higher spatial resolution IFU data with Gemini (GN-2015A-Q-75) of the central region of these galaxies to confirm they are normal elliptical galaxies and constrain a more sophisticated dynamical model of the system.

%--------------------------------------
\acknowledgments
Thanks to Dave Doss and the staff at McDonald Observatory for their help on these observing runs. Thanks to Dmitry Bizyaev for stimulating conversations about GLSBs. Thanks to Lauren MacArthur for discussions on stellar metallicity gradients.

@@

% LocalWords:  Debattista Horrocks Lancashire RCUK Malin Stanonik Spavone Lelli
% LocalWords:  Bothun GLSB Das Mapelli Weij Loebman SungWon Kwak Scd Sd
% LocalWords:  Schiminovich Iodice Belokurov Fairchild RMS pPXF Cappellari th
% LocalWords:  GANDALF Sarzi CosmoloPy Bernardi GLSBs DeBlok Verheijen Yoachim
% LocalWords:  SEDs Brookhaven Instituto Astrofisica Canarias Notre JINA DM
% LocalWords:  IFU UGC HSB SAURON ellipticals Spitzer IC Galaz Myr
% LocalWords:  kpc HIPASS CCD metallicity MANGA SDSS arcsec lensing IRAF
% LocalWords:  Spectrograph spectrograph IDL longslit photometric FWHM SSP mis
% LocalWords:  dispersions Balmer decrement fiducia redshift Chabrier
% LocalWords:  Faber redshifts NGC kinemetry photometry anisotropy Gyr SFH
% LocalWords:  metallicities colliders rotators et al MNRAS de datacube Ppxf
% LocalWords:  dashed CDM Kasparova LSB Dutton Raskutti Sa Morganti
%%  LocalWords:  Emonts Weijmans Ibata McConnachie Schweizer Magain
%%  LocalWords:  Oosterloo baryonic Buote GALEX Salim Kelson outlier
%%  LocalWords:  Bruzual Ahn Willett SQL Chilingarian Saulder Swaters
%%  LocalWords:  Krajn GN Kuntschner situ indices Emsell Capp CasJobs
%%  LocalWords:  Schmitz circ accrete jabref CMD collisional GLSB's
%%  LocalWords:  impactor Conroy brightnesses Malmquist dex Hagen
%%  LocalWords:  Dmitry isothermal

>>>>

\begin{abstract}
We present IFS observations of the giant low surface brightness galaxies Malin~2 and UGC~6614.
We fit the stellar dynamics of these systems and compare their spatially resolved spectra to star formation history models.
We argue that Malin~2 and UGC~6614 are early-type elliptical galaxies that have undergone high angular momentum mergers which formed their GLSB disks, making them unique systems for probing the dark matter halos of elliptical systems.
We find the stellar dynamics are consistent with earlier measurements of the GLSB \HI\ disks.
The stellar populations of the GLSB disks are dominated by old metal-poor stars, while the central ellipticals are metal rich.

@@

Observers tend to classify GLSB systems as late-type spirals.
\citet{Matthews01b} list Malin~2 as an Scd while \citet{Das13} conclude that GLSB systems are an extreme form of late type spiral galaxies.
Similarly, \citet{Kasparova14} claim that the large dark matter scale of Malin~2 can explain it's features.
Even the Galaxy Zoo project \citep{Willett13}  classifies GLSB galaxies as spirals.
While GLSB systems superficially resemble late-type spiral galaxies, they are clearly extreme outliers when their scale lengths are compared to other galaxy samples, being around a factor of 2 larger than the largest spirals \citep[Figure~4]{Dalcanton97}.

@@

The other popular formation scenario for GLSB systems is that they are galaxies that undergo a high angular momentum minor merger where the LSB disk is the shredded remains of an accreted galaxy.
\citet{Pen06} find that extended disks can be formed via the disruption of dwarf galaxies.
These simulated debris disks have exponential profiles and extend out to 200 kpc from the central galaxy.

Observations of Malin~1 support the accretion model of GLSBs.
\citet{Barth07} analyzes HST imaging of Malin~1 and find that the central object is well described as an early-type barred galaxy that was previously unresolved in ground-based imaging. \citet{Barth07} points out that extended disk structures in galaxies such as M31 appear to be a smooth continuation of the inner disk \citep{Ferguson02}, while Malin~1's extended disk is better described as a distinct component, (although \citet{Ibata05} claim M31's extended disk is of accretion origin).

@@

We obtained spatially resolved spectroscopy of our target galaxies using the Mitchell Spectrograph (formerly VIRUS-P)\citep{Hill08} on the 2.7m Harlan J. Smith telescope at McDonald Observatory.
Observations were made during dark time in February, March, and April 2010.
Identical instrument setups were used for all observations.
The Mitchell Spectrograph with the VP-2 IFS (Integral Field Spectrograph) bundle used here has a square array of 246 optical fibers which sample the 1.9\arcmin$\times$1.9\arcmin\ field-of-view with a 1/3 filling factor, with each fiber having a diameter of 4.23\arcsec.
The spectrograph images the spectrum of each fiber on a 2048$\times$2048 Fairchild Imaging CCD.

@@

Once the Vaccine routines are completed we make the following additional reductions using standard IRAF routines and custom IDL scripts.
The reduced twilight frames are used to generate a fiber-to-fiber flux correction (similar to a longslit illumination correction).
We use the b-spline procedure described in \citet{Kelson03} to generate background frames.
The corners of the IFS array contain enough empty fibers to measure the sky background in each exposure.
Because we bin fibers together, imperfect sky-subtraction represents a serious source of potential systematic errors.

@@

\begin{figure*}
\epsscale{.4}
\plotone{Color_Magnitu_Diagram.eps}\plotone{Faber_Jackson_Relation.eps}\\
\plotone{Fundamental_Plane.eps}\plotone{mg_sigma.eps}
\epsscale{1}
\caption{
Top Left:  The color-magnitude diagram for 141,000 galaxies of similar redshift and magnitude as Malin~2 and UGC~6614.
Top Right:  The Faber-Jackson relation for 45,700 SDSS galaxies, along with a best-fit line.
Left:  The $r$-band fundamental plane from SDSS along with our GLSB galaxies and best-fit line.
Bottom Right:  The relation between central velocity dispersion and the Mg2 absorption feature.
 From the point of view of the SDSS data reduction pipeline, GLSB galaxies appear as relatively red elliptical galaxies on the red-sequence.
\label{elliptical_relations}}
\end{figure*}

@@

The diameters of GLSB disks are very large, with Malin~1's stellar disk detected out to a radius of 80 kpc \citep{Galaz15}.
Stellar disks are typically embedded in much larger \HI\ disks.
In Figure~\ref{disk_comp}, we compare the sizes of the stellar and \HI\ disks of the GLSB galaxies with other observations.
For this plot, we take the maximum detected radius from the literature for the galaxies.
It is possible that deeper observations could increase the values of $R_{max}$.
While the stellar disk of UGC~6614 is close in size to local observed \HI\ disks, Malin~1 and~2 are larger than the \HI\ disks observed in the THINGS survey.
This seems to point against the GLSB formation mechanism of \citet{Mapelli08}, as we do not observe \HI\ disks in late-type spirals large enough to convert into GLSB systems unless an interaction is able to also boost the angular momentum of the resulting stellar disk.

@@

As we have shown in \S\ref{sec:centers} and \S\ref{sec:disks}, GLSB galaxies can be well described as massive early-type galaxies that have managed to accrete a high angular momentum disk of stars and gas.
Given the total \HI\ mass in GLSBs, the accreted object is consistent with a much lower-mass gas-rich spiral.
If this is in fact the way GLSBs form, the disk kinematics represent a unique probe of the dark matter halo of elliptical galaxies.

@@

\citet{Murphy2011} are able to use IFS observations of M87 to reach $\sim2 R_e$ using just stellar data, and then rely on globular cluster kinematics to extended out to $5 R_e$ where they find a dark matter fraction of 85\%.
\citet{Weij09} is similarly able to use IFS observations to reach $\sim 4 R_e$ on two elliptical galaxies while \citet{Proctor09} reaches $\sim 3 R_e$ using multi-object spectroscopy.

@@

We find that the centers of these two GLSB systems very closely resemble typical early-type galaxies, with the extended faint disks being distinct components, consistent with the previous results for Malin~1 \citep{Barth07} and NGC~7589 \citep{Lelli10}.

@@

It also seems unlikely that GLSBs are just extreme versions of late-type disk galaxies \citep{Das13}.
It's hard to imagine how such a large massive central spheroid could be formed without also disrupting the surrounding disk.
While the old stellar populations in the disks of Malin~2 and UGC~6614 are evidence against the \citet{Mapelli08} formation scenario, we have not addressed what collisional ring galaxies do evolve into.
Since photometric surveys are often surface brightness limited, it is possible a GLSB system will be found in the future that is the descendant of a collisional ring system.

One of the most important features we can see in our data is the sharp metallicity gradient as one transitions between the central region and the disk in both Malin~2 and UGC~6614 (Figures~\ref{metal_malin2} and~\ref{metal_ugc}).
For both galaxies, the best fitting star formation histories show the stellar populations are dominated by old stars inside and outside the transition, but the metallicity changes by $\sim0.7-1.0$ dex.
This is a very sharp gradient compared to what has been found in late-type spirals for the bulge-to-disk transition region.
\citet{MacArthur09} spectroscopically observed a range of spiral galaxies and found the metallicity profiles which were dropping, rising, and flat.
Using broadband colors, \citet{MacArthur04} also found metallicity gradients were generally very shallow, with steep metallicity gradients only in very late-type spiral galaxies. Also note that our metallicity gradient could be an underestimate of the true gradient since the central metallicities are at the limits of our model grid (i.e., they could have metallicities greater than [Z/H]=0.7). We have also spatially binned data, possibly diluting how sharp the transition is between bulge and disk.

@@

We have observed two giant low surface brightness galaxies with the VIRUS-P IFS.
We find the systems are best explained as massive elliptical galaxies which have managed to accrete stars and gas into a large disk.
We find the inner bulge is similar to elliptical galaxies found in SDSS.
The extended galaxy stars have disk-like velocity fields, are dominated by old stars, but do show signs of ongoing star formation. The stars also show a sharp metallicity transition, consistent with the disk forming independently of the bulge. The \HI\ mass of the disk is also consistent with coming from a low-mass accreted galaxy.

We are currently observing higher spatial resolution IFS data with Gemini (GN-2015A-Q-75) of the central region of these galaxies to confirm they are typical early-type galaxies and constrain a more sophisticated dynamical model of the system.

%--------------------------------------
\acknowledgments
Thanks to Dave Doss and the staff at McDonald Observatory for their help on these observing runs. Thanks to Dmitry Bizyaev for stimulating conversations about GLSBs. Thanks to Lauren MacArthur for discussions on stellar metallicity gradients. Thanks to the anonymous referee for a very thorough and prompt report which improved the paper.

@@

% LocalWords:  Debattista Horrocks Lancashire RCUK Malin Stanonik Spavone Lelli
% LocalWords:  Bothun GLSB Das Mapelli Weij Loebman SungWon Kwak Scd Sd
% LocalWords:  Schiminovich Iodice Belokurov Fairchild RMS pPXF Cappellari th
% LocalWords:  GANDALF Sarzi CosmoloPy Bernardi GLSBs DeBlok Verheijen Yoachim
% LocalWords:  SEDs Brookhaven Instituto Astrofisica Canarias Notre JINA DM
% LocalWords:  IFS UGC HSB SAURON ellipticals Spitzer IC Galaz Myr
% LocalWords:  kpc HIPASS CCD metallicity MANGA SDSS arcsec lensing IRAF
% LocalWords:  Spectrograph spectrograph IDL longslit photometric FWHM SSP mis
% LocalWords:  dispersions Balmer decrement fiducia redshift Chabrier
% LocalWords:  Faber redshifts NGC kinemetry photometry anisotropy Gyr SFH
% LocalWords:  metallicities colliders rotators et al MNRAS de datacube Ppxf
% LocalWords:  dashed CDM Kasparova LSB Dutton Raskutti Sa Morganti
%%  LocalWords:  Emonts Weijmans Ibata McConnachie Schweizer Magain
%%  LocalWords:  Oosterloo baryonic Buote GALEX Salim Kelson outlier
%%  LocalWords:  Bruzual Ahn Willett SQL Chilingarian Saulder Swaters
%%  LocalWords:  Krajn GN Kuntschner situ indices Emsell Capp CasJobs
%%  LocalWords:  Schmitz circ accrete jabref CMD collisional GLSB's
%%  LocalWords:  impactor Conroy brightnesses Malmquist dex Hagen
%%  LocalWords:  Dmitry isothermal
